Formal Analysis of Rate-Monotonic Scheduling Implementation via Real-Time Maude

### 2015-10-01 RMS/submission.tex
finishing section 4

\subsection{Specification}
A Real-Time Maude module specifies a \emph{real-time rewrite theory}
$(\Sigma, E\cup A , IR, TR)$, where:
\begin{itemize}
\item $\Sigma$ is an algebraic \emph{signature}, that is, a set of
  declarations of \emph{sorts}, \emph{subsorts} and \emph{function
    symbols}.
\item $(\Sigma, E\cup A)$ is a \emph{membership equational logic
  theory}, with $E$ a set of possibly conditional equations, and $A$ a
  set of equational axioms such as associativiy, commutativity and
  identity.  $(\Sigma, E\cup A)$ specifies the system's state space as
  an algebraic data type, and includes a built-in specification of a
  sort \verb|Time|.
\item $IR$ is a set of \emph{labeled conditional rewrite rules}
  specifying the system's local transitions, each of which has the
  form $[l]~:~t\rightarrow t'\mbox{ \textbf{if}
  }\bigwedge^n_{j=1}cond_j$, where each $cond_j$ is an equality
  $u_j=v_j$, and $l$ is a \emph{label}. Such a rule specifies an
  \emph{instantaneous transition} from an instance of $t$ to the
  corresponding instance of $t'$, \emph{provided} the conditions hold.
\item $TR$ is a set of \emph{tick rules} $[l]~:~\{t\}\rightarrow\{t'\}
  \mbox{ \textbf{in time} }\tau\mbox{ \textbf{if} }cond$ that advance
  time in the \emph{entire} state $t$ by $\tau$ time units.
\end{itemize}

@@

@@

\subsection{Interrupt Requesting}
Interrupt requests are performed by the source exactly every cycle
$T$, when the attribute \verb|val| decreases to \verb|zero| while time
advances. This is modeled by the following instanous conditional rule
applied on \verb|System|:
\begin{verbatim}
  crl [interrupt-request] :
    (L T STS HW ISRC) => (L T STS (HW).intReq reset(ISRC))
    if (ISRC).timeout .
\end{verbatim}
where the operator \verb|_.timeout| examines whether the attribute
\verb|val| equals to \verb|zero|, and \verb|_.intReq| sets the
\verb|ir| flag indicating there is an interrupt request to be handled:
\begin{verbatim}
  op _.intReq : Hardware -> Hardware .
  eq [ REGS ; S ].intReq = [ (REGS).setIr ; S ] .
\end{verbatim}
Then the request will wait to be handled, which is explained in
Section~\ref{ss:inthandling}.

@@

\subsection{Interrupt Handling and Task Scheduling}
\label{ss:inthandling}
When an interrupt request comes, it may not be ``seen'' immediately by
the system. It requires the \verb|mask| flag to be cleared. Once the
request is seen, it is handled by performing two stages: the interrupt
mechanism of the hardware (such as clearing \verb|ir|, pushing context
into stack and so on) and running the $schedule()$ function. This is
modeled by the following instanous rewrite rule:
\begin{verbatim}
  crl [interrupt-handle] :
    SYSTEM => ((SYSTEM).interrupt).startScheduling
    if (SYSTEM).existInt .
\end{verbatim}
where \verb|_.existInt| checks whether \verb|mask| is cleared
\emph{and} \verb|ir| is set. The operation \verb|_.interrupt| models
the interrupt mechanism performed by the hardware and does four
things: (i)clearing the \verb|ir| flag, which means the request has
been handled; (ii)pushing the current \verb|pc| value into the stack,
storing the interrupted context; (iii)assigning \verb|scheduling| of
sort \verb|Oid| to \verb|pc|, which indicates that $schedule()$ is
running; and (iv)setting the \verb|mask| flag, to mask the coming
interrupt requests. As for \verb|_.startScheduling|, it performs
updating the status of $taskList$ and increasing $timer$ by $1$,
corresponding to Line~\ref{l:updatestatus} and \ref{l:inc} in
Algorithm~\ref{a:schedule}.
\begin{verbatim}
  op _.startScheduling : System -> System .
  eq (L T STS HW ISRC).startScheduling 
       = ((updateStatus L with T) inc(T) STS HW ISRC) .
\end{verbatim}

@@

\subsection{Timed Behaviours of the System}

>>>>

@@

\subsection{Specification}
A Real-Time Maude module specifies a \emph{real-time rewrite theory}
$(\Sigma, E\cup A , IR, TR)$, where:
\begin{itemize}
\item $\Sigma$ is an algebraic \emph{signature}, that is, a set of
  declarations of \emph{sorts}, \emph{subsorts} and \emph{function
    symbols}.
\item $(\Sigma, E\cup A)$ is a \emph{membership equational logic
  theory}, with $E$ a set of possibly conditional equations, and $A$ a
  set of equational axioms such as associativiy, commutativity and
  identity.  $(\Sigma, E\cup A)$ specifies the system's state space as
  an algebraic data type, and includes a built-in specification of a
  sort \verb|Time|.
\item $IR$ is a set of \emph{labeled conditional rewrite rules}
  specifying the system's local transitions, each of which has the
  form $[l]~:~t\rightarrow t'\mbox{ \textbf{if}
  }\bigwedge^n_{j=1}cond_j$, where each $cond_j$ is an equality
  $u_j=v_j$, and $l$ is a \emph{label}. Such a rule specifies an
  \emph{instaneous transition} from an instance of $t$ to the
  corresponding instance of $t'$, \emph{provided} the conditions hold.
\item $TR$ is a set of \emph{tick rules} $[l]~:~\{t\}\rightarrow\{t'\}
  \mbox{ \textbf{in time} }\tau\mbox{ \textbf{if} }cond$ that advance
  time in the \emph{entire} state $t$ by $\tau$ time units.
\end{itemize}

@@

\subsection{Interrupt Requesting}
Interrupt requests are performed by the source exactly every cycle
$T$, when the attribute \verb|val| decreases to \verb|zero| while time
advances. This is modeled by the following instaneous conditional rule
applied on \verb|System|:
\begin{verbatim}
  crl [interrupt-request] :
    (L T STS HW ISRC) => (L T STS (HW).intReq reset(ISRC))
    if (ISRC).timeout .
\end{verbatim}
where the operator \verb|_.timeout| examines whether the attribute
\verb|val| equals to \verb|zero|, and \verb|_.intReq| sets the
\verb|ir| flag indicating there is an interrupt request to be handled:
\begin{verbatim}
  op _.intReq : Hardware -> Hardware .
  eq [ REGS ; S ].intReq = [ (REGS).setIr ; S ] .
\end{verbatim}
Then the request will wait to be handled, which is explained in
Section~\ref{ss:inthandling}.

@@

\subsection{Interrupt Handling and Task Scheduling}
\label{ss:inthandling}
When an interrupt request comes, it may not be ``seen'' immediately by
the system. It requires the \verb|mask| flag to be cleared. Once the
request is seen, it is handled by performing two stages: the interrupt
mechanism of the hardware (such as clearing \verb|ir|, pushing context
into stack and so on) and running the $schedule()$ function. This is
modeled by the following instaneous rewrite rule:
\begin{verbatim}
  crl [interrupt-handle] :
    SYSTEM => ((SYSTEM).interrupt).startScheduling
    if (SYSTEM).existInt .
\end{verbatim}
where \verb|_.existInt| checks whether \verb|mask| is cleared
\emph{and} \verb|ir| is set. The operation \verb|_.interrupt| models
the interrupt mechanism performed by the hardware and does four
things: (i)clearing the \verb|ir| flag, which means the request has
been handled; (ii)pushing the current \verb|pc| value into the stack,
storing the interrupted context; (iii)assigning \verb|scheduling| of
sort \verb|Oid| to \verb|pc|, which indicates that $schedule()$ is
running; and (iv)setting the \verb|mask| flag, to mask the coming
interrupt requests. As for \verb|_.startScheduling|, it performs
updating the status of $taskList$ and increasing $timer$ by $1$,
corresponding to Line~\ref{l:updatestatus} and \ref{l:inc} in
Algorithm~\ref{a:schedule}.
\begin{verbatim}
  op _.startScheduling : System -> System .
  eq (L T STS HW ISRC).startScheduling 
       = ((updateStatus L with T) inc(T) STS HW ISRC) .
\end{verbatim}

@@

\subsection{Timed Behaviours of the System}
\emph{Timed behaviours} of the system consist of two parts, the
running of tasks, including both periodic ones and system ones, and
the running of interrupt source. These are all modeled as the
following one conventional tick rule:
\begin{verbatim}
  crl [tick]:
    {SYSTEM} => {delta(SYSTEM, R)} in time R 
    if R le mte(SYSTEM) [nonexec] .
\end{verbatim}
where \verb|delta| defines the effects of time elapse on the system,
and \verb|mte| denotes the maximum amount of time allowed to elapse
from the current state until an instaneous transition \emph{must} be
performed. Notice that the variable $R$ is \emph{continuous} with
respect to the specific time domain\footnote{Real-Time Maude contains
  built-in modules to define the time domain to be natural numbers and
  rational numbers, specifying \emph{discrete} time domains and
  \emph{dense} time domains respectively.}  that we choose to
instantiate the model on, which is different from timed automata that
discretize dense time by defining ``clock region''.

Time affects the system by advancing both the running task, whose $ID$
is loaded at \verb|pc|, and the interrupt source simultaneously,
increasing their \verb|cnt| and \verb|val| respectively:
\begin{verbatim}
  ceq delta((L T STS HW ISRC), R)
        = ((deltaTask(ID, L, R) T STS HW (deltaIS(ISRC, R)))
      if ID := (HW).getPc /\ ID :: MaybeNat .
\end{verbatim}
We ease the reading by not showing the case where $ID$ is of sort
\verb|Oid|, due to similarity. At that case, \verb|deltaTask| applies
on \verb|STS| instead of \verb|L|.

The maximum amount of time allowed to elapse depends on three aspects:
the time remained to finish the running task, either periodic one or
system one, the time remained to request an interrupt, and whether or
not there exists a non-masked interrupt request for the moment:
\begin{verbatim}
  ceq mte(L T STS HW ISRC)
        = minimum( mteTask(ID, L),
                   minimum( mteIS(ISRC), mteIr(HW)))
      if ID := (HW).getPc /\ ID :: MaybeNat .
\end{verbatim}
where \verb|mteIr| will return \verb|zero| if there exists a
non-masked interrupt request in the system, or \verb|INF| which
represents \emph{infinity} otherwise. Again we do not show the case
where $ID$ is of sort \verb|Oid|, which is very similar. We should
point out that \verb|mteTask| computes the time remained to reach the
maximum of the possible finishing interval of the task, since it is
the time until which a ``finish'' transition \emph{must} happen.
